An iterated Grothendieck construction of semi-cubical types

### 2021-09-08 paper/paper.tex
paper: More notes on formalization; arrows not rendering in minted

% The eqntable environment, displaying the various
\NewDocumentEnvironment{eqntable}{m}{\table[!ht]
\tabularx{\textwidth}{@{} l c c >{\centering\arraybackslash}X @{}}\toprule}
{\endtabularx\hrule\caption{#1}\endtable}

@@

\begin{document}
\begin{abstract}
  A fresh approach to formalizing semi-cubical sets in dependent type theory with extensional equality. The approach could be described as a \emph{reverse iterated Grothendieck construction}, that is crisp and concise. The accompanying artifact is a Coq formalization that uses highly sophisticated rewriting techniques and pushes the boundary of proof assistant technology.
\end{abstract}
\maketitle
\tableofcontents
\newpage

\section{Overture}
Cubical type theory is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}. Cubical type theory has several attractive features, and people have been working on it since 2014~\cite{Bezem14}. Today, an experimental branch of Agda, known as \emph{Cubical Agda} is being worked on~\cite{Vezzosi21}.

However, there is no known formalization of cubical types in \emph{Coq}, which is a vastly more powerful proof assistant than Agda. Our work lays the foundation for a future Cubical Coq.

\section{The intutution from simplicial sets}
Simplicial sets are powerful mathematical objects, that form the basis of much of modern homotopy theory. Their power derives from the fact that their homotopy category is exceptionally well-behaved, and different model structures on them, as we will see in the following chapter. Unlike CW-complexes that rely on pure topological notions like spheres and disks to ``tame'' a topological space, the definition of simplicial sets is purely categorical, and exist independent of topological spaces. The geometric realization and singular complex functors facilitate travelling back-and-forth between simplicial sets and topological spaces, and these are non-trivial constructions. As such, simplicial sets exist purely in the imagination of the mathematician; each simplicial set has an infinite number of degenerate simplices, and they are not particularly well-suited to machine-based representation.

@@

\section{The formalization effort}
\subsection{Converging on an approach}
The core of our formalization effort demands proof irrelevance for inequalities. There were two approaches on the outset: one was HoTT, where we'd have proof relevance, but be able to transport one proof to another, and another was SProp, where we'd get proof irrelevance out-of-the-box. We initially experimented with HoTT, but there were serious limitations: HoTT drops the Coq standard library, and several tactics are (unfortunately) reliant on the standard library. As a result, several sophisticated tactics are infeasible in HoTT.

SProp comes with a host of bugs and it took us some time to become comfortable with it, and realize its limitations. We did go back and forth with SProp a bit, but ended up using it in the final version.

\subsection{Yoneda.v}
\subsection{RewLemmas.v}
\subsection{Notes on SProp}
\subsection{Cubical.v}

We'd tried to build the \texttt{Cubical} as a \texttt{Fixpoint} initially, but ran into unification problems with \texttt{le\_dec}.

Type unification on the term \texttt{D} fails.

\begin{minted}{coq}
From Coq Require Import Arith.

Theorem le_S_up {n m} (Hnm : n <= m) : n <= S m.
Admitted.
Theorem le_S_down {n m} (Hnm : S n <= m) : n <= m.
Admitted.
Theorem le_discr {n} : S n <= 0 -> forall {A}, A.
Admitted.
Theorem propagate_eq {m n} : m <= S n -> m = S n -> n <= n.
Admitted.
Theorem le_dec {n m} : n <= S m -> {n = S m} + {n <= m}.
Admitted.
Theorem le_dec' {n m} : n <= m -> {n = m} + {n <= pred m} + {m = O}.
Admitted.
Theorem lower_both {m n} : S m <= S n -> m <= n.
Admitted.

Notation "⇓ p" := (le_S_down p) (at level 40).
Notation "↑ h" := (le_S_up h) (at level 40).
Notation "l '.1'" := (projT1 l) (at level 40).
Notation "l '.2'" := (projT2 l) (at level 40).

Section Cubical.

Universe l'.
Universe l.

Record Cubical {n : nat} :=
{
csp {n'} (Hn' : n' <= n) : Type@{l'} ;
hd {n'} {Hn' : S n' <= n} : csp Hn' -> csp (⇓ Hn') ;
box {n' p} {Hn' : n' <= n} (Hp : p <= n') : csp Hn' -> Type@{l} ;
}.

Fixpoint cubical {n : nat} : Cubical :=
match n with
| O => {| csp _ _ := unit;
hd _ Hn' _ := ltac:(apply (le_discr Hn'));
box _ _ _ _ _ := unit;
|}
| S n => let cn := cubical (n := n) in
let cspn {n'} (Hn' : n' <= S n) :=
match le_dec Hn' with
| left Heq (* n' = S n *) => { D : cn.(csp) (propagate_eq Hn' Heq) &
cn.(box) _ D -> Type@{l} }
| right Hineq (* n' <= n *) => cn.(csp) Hineq
end in
let hdn {n'} (Hn' : S n' <= S n) :=
match le_dec' (lower_both Hn') return cspn Hn' (* S n' <= S n *) ->
cspn (⇓ Hn') (* n' <= S n *) with
| inleft Hcmp =>
match Hcmp with
| left _ (* n' = n *) => fun D => D.1 (* n' = S n *)
| right _ (* n' <= pred n *) => fun D => cn.(hd) D (* n' <= n *)
end
| inright _ (* n = O *) => fun D => cn.(hd) D
end in
  {|
    csp _ Hn' := cspn Hn';
    hd _ Hn' := hdn Hn';
    |}
end.

End Section.
\end{minted}

>>>>

% The eqntable environment, displaying the various
\NewDocumentEnvironment{eqntable}{m}{\table[!ht]\small
\tabularx{\textwidth}{@{} l c c >{\centering\arraybackslash}X @{}}\toprule}
{\endtabularx\vspace{0.3em}\hrule\vspace{0.5em}\caption{#1}\endtable}

@@

\begin{document}
\begin{abstract}
  A proposed model of \emph{homotopy type theory} built via a \emph{reverse iterated Grothendieck construction} process formalizing semi-cubical sets in dependent type theory with \emph{extensional equality}. The accompanying artifact is a Coq formalization that uses highly sophisticated rewriting techniques and pushes the boundary of proof assistant technology.
\end{abstract}
\maketitle
\tableofcontents
\newpage

\section{Overture}
Cubical type theory is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. Cubical type theory has several attractive features, and people have been working on it since 2014~\cite{Bezem14}. Today, an experimental branch of Agda, known as \emph{Cubical Agda} is being worked on~\cite{Vezzosi21}.

Our contribution takes a different direction from traditional contributions to the field of HoTT: although it is a proposed model of HoTT, it uses \emph{unicity of identity proofs} (UIP) instead of univalence to finish off the final coherence condition; the two axoims are mutually incompatible.

\section{The intuition from simplicial sets}
Simplicial sets are powerful mathematical objects, that form the basis of much of modern homotopy theory. Their power derives from the fact that their homotopy category is exceptionally well-behaved, and different model structures on them, as we will see in the following chapter. Unlike CW-complexes that rely on pure topological notions like spheres and disks to ``tame'' a topological space, the definition of simplicial sets is purely categorical, and exist independent of topological spaces. The geometric realization and singular complex functors facilitate travelling back-and-forth between simplicial sets and topological spaces, and these are non-trivial constructions. As such, simplicial sets exist purely in the imagination of the mathematician; each simplicial set has an infinite number of degenerate simplices, and they are not particularly well-suited to machine-based representation.

@@

\section{The formalization effort}
The core of our formalization effort demanded proof irrelevance for inequalities: \mintinline{coq}{SProp} was the natural choice.

\subsection{\texorpdfstring{\leq}{<=} in SProp}
We start with one inequality primitive, namely $\leq$, and build up all our inequalities on the basis of that inequality. We tried three different approaches to defining $\leq$ in \mintinline{coq}{SProp}.

\begin{enumerate}
  \item[(a)] An inductive definition.
    \begin{listing}[H]
      \begin{minted}{coq}
      Inductive le' (n : nat) : nat -> SProp :=
      | le_refl : n <= n
      | le_S_up : forall {m : nat}, n <= m -> n <= S m
      where "n <= m" := (le' n m) : nat_scope.

    Definition le := le'.
    \end{minted}
    \end{listing}
  \item[(b)] A recursive definition.
    \begin{listing}[H]
      \begin{minted}{coq}
      Inductive sFalse : SProp :=.
      Inductive sTrue : SProp := sI.

      Fixpoint le m n : SProp :=
        match m with
        | 0 => sTrue
        | S m =>
          match n with
          | 0 => sFalse
          | S n => le m n
          end
        end.
     \end{minted}
    \end{listing}
  \item[(c)] An nested implication.
    \begin{listing}[H]
      \begin{minted}{coq}
      Inductive le' (n : nat) : nat -> SProp :=
        | le_refl' : n <= n
        | le_S_up' : forall (m : nat), n <= m -> n <= S m
        where "n <= m" := (le' n m) : nat_scope.

      Definition le n m := forall p, p <= n -> p <= m.
      \end{minted}
    \end{listing}
\end{enumerate}

(b) was unworkable right from the start, and we tried approach (a) for a while, before settling on approach (c), because it made compositions of inequalities almost trivial to prove. Compare the following snippets defining transitivity of $\leq$:

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Theorem le_trans {p q n} : p <= q -> q <= n -> p <= n.
    (* ... a non-trivial proof ... *)
    Defined.
  \end{minted}
  \caption{↕ definition, when $\leq$ is defined using an inductive approach}
\end{listing}

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Definition le_trans {n m p} (Hnm : n <= m) (Hmp : m <= p) : n <= p :=
      fun q (Hqn : le' q n) => Hmp _ (Hnm _ Hqn).
  \end{minted}
  \caption{↕ definition, when $\leq$ is defined using a nested implication}
\end{listing}

Implications are naturally easier to work with, so the nested implication in \mintinline{coq}{SProp} was an easy choice. Many of the commutativity proofs were a simple \mintinline{coq}{reflexivity}:

\begin{listing}[H]
  \begin{minted}{coq}
  Theorem le_trans_comm4 {n m p} (Hmn : S n <= S m) (Hmp : S m <= S p) :
  ⇓ (Hmn ↕ Hmp) =S (⇓ Hmn) ↕ (⇓ Hmp).
    reflexivity.
  Defined.
  \end{minted}
\end{listing}

\subsection{Raising and lowering operations on \texorpdfstring{\leq}{<=}}
\mint{coq}/Notation "↕" := (Hnm : n <= m) (Hmp : m <= p) : n <= p/
\mint{coq}/Notation "↑" := (Hnm : n <= m) : n <= S m/
\mint{coq}/Notation "↓" := (Hnm : S n <= m) : n <= m/
\mint{coq}/Notation "⇓" := (Hnm : S n <= S m) : n <= m/
\mint{coq}/Notation "⇑" := (Hnm : n <= m) : S n <= S m/
\subsection{Yoneda.v}
\subsection{RewLemmas.v}
\subsection{Notes on SProp}
\subsection{Cubical.v}

We'd tried to build the \mintinline{coq}{Cubical} as a \mintinline{coq}{Fixpoint} initially, but ran into unification problems with \mintinline{coq}{le_dec}.

Type unification on the term \mintinline{coq}{D} fails.

\begin{listing}[H]
  \begin{minted}[linenos]{coq}
    Theorem le_dec {n m} : n <= S m -> {n = S m} + {n <= m}.
    Admitted.
    Theorem le_dec' {n m} : n <= m -> {n = m} + {n <= pred m} + {m = O}.
    Admitted.
    (* ... *)
    let hdn {n'} (Hn' : S n' <= S n) :=
    match le_dec' (lower_both Hn') return cspn Hn' (* S n' <= S n *) ->
    cspn (⇓ Hn') (* n' <= S n *) with
    | inleft Hcmp =>
    match Hcmp with
    | left _ (* n' = n *) => fun D => D.1 (* n' = S n *)
    | right _ (* n' <= pred n *) => fun D => cn.(hd) D (* n' <= n *)
    end
    (* ... *)
    \end{minted}
\end{listing}
